Meeting the Challenges of Modeling Astrophysical Thermonuclear Explosions: \castro, \maestro, and the \amrex\ Astrophysics Suite

### 2017-10-02 astronum_2017/astronum-amrex.tex
more work on this proceedings

@@

The largest open theoretical question for SNe Ia is---what is the
nature of the progenitor?  About 20 years ago, the community had
mostly converged up the single-degenerate scenario---a
Chandrasekhar-mass C/O white dwarf that accretes from its companion,
eventually leading to runaway at the center that burns through the
star (see~\cite{hillebrandtniemeyer2000} for the state of the field
at that time).  Since then, the wealth of observations has show that
there is a lot of diversity in SNe Ia, and searches for progenitor
systems have not shown that Chandra-mass white dwarfs can explain all
SNe Ia.  Today, merging white dwarfs (the double degenerate scenario)
have become the most popular model.  Other progenitors, like He
burning on the surface of a sub-Chandra white dwarf have also seen
interest in explaining some of the observed diversity.

There are open questions in all of these scenarios which can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots setup
by turbulent convection that give rise to burning fronts.  Can you
create a detonation in the He layer in the double detonation model?
For double degenerates, does the burning take place promptly?  Is it
possible to accurate model a detonation in this case, given our
resolution constraints?  These are some of the questions we seek to
answer.

@@

X-ray bursts, the burning of accreted H/He on the surface of a neutron
star, can be important probes of neutron star structure.  Interpreting
observations requests that we understand what we are seeing, which can
be influenced by the products of the burning and how the burning
spreads across the star.  Many efforts have focused on different
aspects of these events.  One-dimensional models capture the
energetics well and inform us about the nucleosynthesis
\cite{woosley-xrb}.  Global models show the importance of rotation in
confining the burning \cite{SPIT_ETAL02}, why models inspired from
atmospheric science can explore the vertical structure
\cite{cavecchi:2012}.  However, the resolution differences from the
scale of the burning to a reasonable fraction of the neutron star
surface has prevented detailed explorations of the burning in resolved
calculations and how it feeds back on the flame structure and
propagation.

@@

conservation of gravity / angular momentum
                strong coupling of hydrodynamics and burning
                long timescale evolution

                3-d, turbulence, instabilities, rotation, ...

@@

Our suite of application codes is built on the
\amrex\ block-structured adaptive mesh refinement (AMR) library.  The
basic programming model is that \amrex\ manages the grid
data-structures and parallelism, and calls the computational kernels
on a patch-by-patch basis.  AMReX supports subcycling in time, which
we use in \castro.

\amrex\ uses a hybrid approach to parallelism.  Tiling is used to
logically subdivide AMR patches to increase the parallelism for
OpenMP, without incurring the memory overheads that using many more
smaller grids would bring~\cite{tiling}.  This strategy is especially
important for many core architectures like the Intel Phi.  Ongoing development
is being done to support GPU offloading, using managed memory provided
by the latest generations of GPUs.

There are many application codes built on \amrex, including those in
combustion, ...  In astrophysics, these include \maestro\ and
\castro\ for stellar and nuclear astrophysics applications and
\nyx~\cite{nyx} for cosmological applications.  Here we focus on the
former two.

@@

\maestro~\cite{MAESTRO:Multilevel} is a low Mach number stellar
hydrodynamics code.  \maestro\ has been applied to convection in the
Chandrasekhar-mass model for Type Ia supernovae (SNe Ia)
\cite{ZABNW:IV,wdconvect,wdturb}, the sub-Chandra model for SNe Ia
\cite{subchandra,subchandra2} X-ray bursts~\cite{xrb,xrb2,xrb3}, and
convection in massive stars \cite{ms_cc}.

\maestro\ decomposes the state into a one-dimensional hydrostatic base
state and a three-dimensional Cartesian state that evolves the
departures from HSE.  A constraint equation is derived by requiring
that the pressure everywhere be close to the background hydrostatic
state.  The constraint acts to enforce instantaneous acoustic
equilibriation, effectively filtering soundwaves from the system,
while retaining the compressibility effects due to the background
stratification of the star and local heat release, as well as the
hydrostatic adjustment of the star.  In this fashion, it is more
general than traditional anelastic methods.  Low speed methods have
seen a lot of recent development~\cite{kleinpauluis,vasil:2013}, which
has been incorporated into \maestro.  The next development effort for
\maestro\ will be to allow for moderate rotation of the star.

@@

\castro~\cite{castro,castroII,castroIII} is a fully-compressible
radiation hydrodynamics code that supports arbitrary equations of
state, nuclear reaction networks, and Poisson gravity.  \castro\ has
been applied to core-collapse supernovae~\cite{castro-ccsne},
population III pair-instability
supernovae~\cite{castro-pairinstability}, and white dwarf mergers as a
model for SNe Ia~\cite{wdmergerI}.  Recent work has focused on
improving energy conservation with gravitational and rotational
sources and new integration methods that more accurately couple hydro
and reactions.

For \maestro\ simulations that evolve from the subsonic regime to the
sonic regime, we have demonstrated the ability to restart the
calculations in \castro\ to continue the evolution into the sonic
regime~\cite{scidac-petascale,malone:2014} (in this case, for Chandra
model SNe Ia).

@@

>>>>

\usepackage{color}
\setlength{\marginparwidth}{0.75in}
\newcommand{\MarginPar}[1]{\marginpar{\vskip-\baselineskip\raggedright\tiny\sffamily\hrule\smallskip{\color{red}#1}\par\smallskip\hrule}}

@@

The largest open theoretical question for SNe Ia is---what is the
nature of the progenitor?  About 20 years ago, the community had
mostly converged up the single-degenerate scenario---a
Chandrasekhar-mass C/O white dwarf that accretes from its companion,
eventually leading to runaway at the center that burns through the
star (see~\cite{hillebrandtniemeyer2000} for the state of the field at
that time).  Since then, the wealth of observations has show that
there is a lot of diversity in SNe Ia, and searches for progenitor
systems have not shown that Chandra-mass white dwarfs can explain most
SNe Ia.  Today, merging white dwarfs (the double degenerate scenario)
have perhaps become the most popular model.  Other progenitors, like
He burning on the surface of a sub-Chandra white dwarf have also seen
interest in explaining some of the observed diversity.
See~\cite{araa-maoz} for a review.

There are open questions in all of these scenarios which can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots setup
by turbulent convection that give rise to burning fronts?  A
longstanding question with the Chandra model is, can a deflagration
transition to a detonation during the burning front propagation
through the star.  For the sub-Chandra double detonation model, is it
possible to create a detonation in the thin surface He layer?  For
double degenerates, does the burning take place promptly? and can we
avoid an accretion-induced collapse to a neutron star.  For many of
these, we have the question of whether it is possible to accurately
model the ignition of a detonation with the spatial resolution we can
attain?  These are some of the questions we seek to answer.

@@

X-ray bursts, the burning of accreted H/He on the surface of a neutron
star, can be important probes of neutron star structure.  Interpreting
observations requires that we understand what we are seeing, which can
be influenced by the products of the burning and how the burning
spreads across the star.  Many efforts have focused on different
aspects of these events.  One-dimensional models capture the
energetics well and inform us about the nucleosynthesis
\cite{woosley-xrb}.  Global models show the importance of rotation in
confining the burning \cite{SPIT_ETAL02}, while models inspired from
atmospheric science can explore the vertical structure
\cite{cavecchi:2012}.  However, the resolution differences from the
scale of the burning to a reasonable fraction of the neutron star
surface has prevented detailed explorations of the burning in resolved
calculations and how it feeds back on the flame structure and
propagation.  Algorithms and computing architectures are starting to
reach the point where we can span the gaps in spatial scales between
these calculations to provide a better understanding of the ignition
and propagation of the burning front, and the nucleosynthesis produced.

@@

These problems share common algorithmic requirements: strong coupling
between hydrodynamics and burning, support for a general equation of state,
self-gravity, including isolated boundary conditions, and long timescale
evolution for the convective phases.  All of these problems are inherently
three-dimensional, as turbulence, fluid instabilities, and rotation affect
the dynamics.  Conservation is important as well, suggesting approaches
which implement gravitational and rotation sources conservatively, and
methods for improving angular momentum conservation.

@@

Our suite of application codes is built on the
\amrex\ block-structured adaptive mesh refinement (AMR) library.  The
basic programming model has \amrex\ managing the grid data-structures
and parallel communications, and it calls the computational kernels on
a patch-by-patch basis.  The core library is written in C++ with the
computational kernels written in Fortran\footnote{Currently
  \maestro\ is using an older pure-Fortran interface, but this will be
  ported to the C++ AMReX this coming year.}---this allows us to take
advantage of the strengths of both languages.AMReX supports subcycling
in time, which we use in \castro.

\amrex\ uses a hybrid MPI + OpenMP approach to parallelism.
Distribution of grid patches to nodes using MPI provides a natural
coarse-grained approach to distributing the computational work, while
threading of loops over zones in a grid using OpenMP provides
effective fine-grained parallelization.  For \castro\, tiling is used
to logically subdivide AMR patches to increase the parallelism for
OpenMP, without incurring the memory overheads that using many more
smaller grids would bring~\cite{tiling}.  This strategy is especially
important for many core architectures like the Intel Phi.  Ongoing
development is being done to support GPU offloading, using managed
memory provided by the latest generations of GPUs.

There are many application codes built on \amrex, including those in
combustion, ... \MarginPar{domains?} In astrophysics, these include \maestro\ and
\castro\ for stellar and nuclear astrophysics applications and
\nyx~\cite{nyx} for cosmological applications.  Here we focus on the
former two.

@@

\maestro~\cite{MAESTRO:Multilevel} is a low Mach number stellar
hydrodynamics code designed for efficiently modeling convection in
stars.  \maestro\ decomposes the state into a one-dimensional
hydrostatic base state and a three-dimensional Cartesian state that
evolves the departures from HSE.  A constraint equation is derived by
requiring that the pressure everywhere be close to the background
hydrostatic state.  The constraint acts to enforce instantaneous
acoustic equilibriation, effectively filtering soundwaves from the
system, while retaining the compressibility effects due to the
background stratification of the star and local heat release, as well
as the hydrostatic adjustment of the star.  In this fashion, it is
more general than traditional anelastic methods.  The timestep
constraint for these equations depends only on the fluid velocity, not
the sound speed, enabling much larger timesteps than compressible
codes for highly subsonic flows.

The state is advanced using a second-order accurate projection method.
Fluid quantities are advected using an unsplit Godunov method, with
reactions incorporated via operator splitting.  The provisional
velocities are then projected onto the space that satisfies the
divergence constraint.  The projections involve elliptic solves,
solving a variable-coefficient Poisson problem, computed using a
multigrid algorithm.  Low speed methods have seen a lot of recent
development~\cite{kleinpauluis,vasil:2013}, which has been
incorporated into \maestro.  The next development effort for
\maestro\ will be to allow for moderate rotation of the star.

\maestro\ has been applied to convection in the Chandrasekhar-mass
model for Type Ia supernovae (SNe Ia)
\cite{ZABNW:IV,wdconvect,wdturb}, the sub-Chandra model for SNe Ia
\cite{subchandra,subchandra2} X-ray bursts~\cite{xrb,xrb2,xrb3}, and
convection in massive stars \cite{ms_cc}.

@@

\castro~\cite{castro,castroII,castroIII} is a fully-compressible
radiation hydrodynamics code that supports arbitrary equations of
state, nuclear reaction networks, and Poisson gravity using geometric
multigrid.  The main hydrodynamics scheme in \castro\ is an unsplit
piecewise parabolic method.  Recent developments have improved
conservation with gravity sources and rotation~\cite{wdmergerI}.  The
radiation solver in \castro\ uses the flux-limited diffusion
approximation for gray or multigroup radiation.  The integration
algorithm on the grid hierarchy is a recursive procedure in which
coarse grids are advanced in time, fine grids are advanced multiple
steps to reach the same time as the coarse grids and the data at
different levels are then synchronized. The synchronization for
self-gravity is similar to the algorithm introduced by
\cite{miniati-colella}.

\castro\ has been applied to core-collapse
supernovae~\cite{castro-ccsne}, population III pair-instability
supernovae~\cite{castro-pairinstability}, and white dwarf mergers as a
model for SNe Ia~\cite{wdmergerI}.  Recent work has focused on
improving energy conservation with gravitational and rotational
sources and new integration methods that more accurately couple hydro
and reactions.  For \maestro\ simulations that evolve from the
subsonic regime to the sonic regime, we have demonstrated the ability
to restart the calculations in \castro\ to continue the evolution into
the sonic regime~\cite{scidac-petascale,malone:2014} (in this case,
for Chandra model SNe Ia).

@@

regression testing
